---
title: 'RNAseq validations DEseq2 analysis'
author: "Florian Wünnemann"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load packages
library(tximport)
library(DESeq2)
library(tidyverse)
library(data.table)
library(plotly)
library(cowplot)
library(ggrepel)
library(GGally)
library(UpSetR)
library(apeglm)
library(here)

theme_set(theme_cowplot())
```

# Introduction

Transcript abundances were quantified using kallisto against ensemble version 96 reference.
Processing of fastq files for RNAseq can be found in:
[rnaseq_validations.create_kallisto_index.sh](./scripts/rnaseq_validations.create_kallisto_index.sh)
[rnaseq_validations.run_kallisto_quant.sh](./scripts/rnaseq_validations.run_kallisto_quant.sh)

# Load metadata

First, we will read in a metadata table that contains the information concerning all experiments.

```{r, eval = TRUE}
metadata <- fread(here("./data/rnaseq.validation_samples_overview.tsv"))
```


# Import counts using tximport

Then, we need to read in our table to aggregate transcript level counts to gene level counts.

```{r}
## Read in tx to gene table
tx2g <- fread(here("./references/tx2gene.with_Cas9.txt"),
              header = FALSE)

tx2g <- tx2g %>%
  select(V1,V3)
```

## All samples together for UMAP and comparison

Now we will use tximport to aggregate all of the samples into one large matrix with counts aggregated at the gene level. This reads the processed data that has been quantified using kallisto.

```{r}
kallisto_dir <- here("./data/kallisto_quant_res")

sample_names <- unique(metadata$sample_name)
files <- paste(kallisto_dir,sample_names,"abundance.h5",sep="/")
names(files) <- sample_names

txi_all <- tximport(files, 
                    type = "kallisto", 
                    txOut = FALSE,
                    tx2gene = tx2g,
                    countsFromAbundance = "scaledTPM")
```

We will then use the tximport data to run DESeq analysis on the different experiments. For this, we will first create a DESeq object.

```{r}
sampleTable <- metadata %>%
  select(sample_name,variant,sgRNA,Cas_flavour,candidate_gene,type,experiment_number,sequencing_facility) %>%
  mutate("experiment_number" = paste("exp",experiment_number,sep="")) %>%
  mutate("experiment_id" = paste(experiment_number,sgRNA,candidate_gene,sep="_")) %>%
  mutate("experiment_id" = as.factor(experiment_id)) %>%
  unique()

rownames(sampleTable) <- sampleTable$sample_name

dds <- DESeqDataSetFromTximport(txi_all, sampleTable, design = ~ experiment_id)
dds <- estimateSizeFactors(dds)
```


```{r}
library("vsn")
keep <- rowSums(counts(dds)) >= 10
dds_filt <- dds[keep,]
rlog_dds <- rlog(dds_filt, blind=TRUE) ## rlog transformation
meanSdPlot(assay(rlog_dds))
```


## PCA based on normalized expression

```{r}
## Check PCA mapping for several metadata factors
plotPCA(rlog_dds, intgroup=c("experiment_number"))
plotPCA(rlog_dds, intgroup=c("sequencing_facility"))
plotPCA(rlog_dds, intgroup=c("Cas_flavour"))
plotPCA(rlog_dds, intgroup=c("experiment_id"))
pcs <- prcomp(assay(rlog_dds))
```

```{r}
## Plot eigenvalues
library(factoextra)
fviz_eig(pcs, addlabels = TRUE, ylim = c(0, 50))

## Find genes driving PC1
pc1_genes <- as.data.frame(pcs$x) %>%
  mutate("gene" = rownames(pcs$x)) %>%
  select(gene,PC1) %>%
  arrange(desc(PC1))
```


```{r}
## Sample distance matrix
sampleDists <- dist(t(assay(rlog_dds)))
```

```{r}
library(RColorBrewer)
library(pheatmap)
sampleDistMatrix <- as.matrix(sampleDists)
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```



```{r}
## Save rlog normalized expression
rlog_dds_df <- as.data.frame(assay(rlog_dds))
rownames(rlog_dds_df) <- rownames(assay(rlog_dds))

## Save raw counts from DESeq2
raw_counts <- as.data.frame(counts(dds))
```

## Compare rlog normalization with scaledTPM values

```{r}
scaledtpm_sub <- txi_all$abundance[rownames(rlog_dds_df),]

merged_exp <- cbind(scaledtpm_sub[,1],rlog_dds_df[,1])
colnames(merged_exp) <- c("scaledTPM","rlog")
merged_exp <- as.data.frame(merged_exp)


ggplot(merged_exp,aes(log2(scaledTPM),rlog)) +
  geom_point() +
  geom_point(data = subset(merged_exp,rownames(merged_exp) %in% c("FES","FURIN","DHX38")),
             color = "red", size = 2) +
    geom_point(data = subset(merged_exp,rownames(merged_exp) %in% c("Cas9")),
             color = "purple", size = 2)
```


## Check Cas9 expression in all experiments

```{r}
cas9 <- rlog_dds_df["Cas9",]
cas9_long <- cas9 %>%
  gather(sample,exp)

ggplot(cas9_long,aes(sample,exp)) +
  geom_bar(stat="identity") +
  coord_flip() +
  labs(y = "Cas9 expression")
```



## Perform DE analysis

```{r}
## Perform differential expression
dds_filt <- DESeq(dds_filt)
```


### Calculate de results

```{r}
## Produce results tables for all experiment comparisons
dds <- dds_filt
dds <- DESeq(dds)
resultsNames(dds)

## FES/FURIN - rs12906125
dds$experiment_id <- relevel(dds$experiment_id, ref="exp1_safe_FES_FURIN")
dds <- DESeq(dds)
experiment_1_furin_fes_lfcshrink <- lfcShrink(dds,
                                              coef = "experiment_id_exp1_sgRNA_06939_FES_FURIN_vs_exp1_safe_FES_FURIN", 
                                              type = 'apeglm')

## ZNF664 - rs12311848
dds$experiment_id <- relevel(dds$experiment_id, ref="exp2_safe_ZNF664_CNNM2")
dds <- DESeq(dds)
experiment_2_znf664_lfcshrink <- lfcShrink(dds,
                                           coef = "experiment_id_exp2_sgRNA_06147_ZNF664_vs_exp2_safe_ZNF664_CNNM2", 
                                           type = 'apeglm')


## CNNM2 - rs78260931
dds$experiment_id <- relevel(dds$experiment_id, ref="exp2_safe_ZNF664_CNNM2")
dds <- DESeq(dds)
experiment_3_cnnm2_lfcshrink <- lfcShrink(dds,
                                           coef = "experiment_id_exp3_sgRNA_25344_CNNM2_vs_exp2_safe_ZNF664_CNNM2", 
                                           type = 'apeglm')

## DHX38 - rs2074626
dds$experiment_id <- relevel(dds$experiment_id, ref="exp4_safe_DHX38")
dds <- DESeq(dds)
experiment_4_dhx38_lfcshrink <- lfcShrink(dds,
                                          coef = "experiment_id_exp4_sgRNA_10966_DHX38_vs_exp4_safe_DHX38", 
                                          type = 'apeglm')


## MAT2A - rs1078004
dds$experiment_id <- relevel(dds$experiment_id, ref="exp5_safe_MAT2A")
dds <- DESeq(dds)
experiment_5_mat2a_lfcshrink <- lfcShrink(dds,
                                          coef = "experiment_id_exp5_sgRNA_02249_MAT2A_vs_exp5_safe_MAT2A", 
                                          type = 'apeglm')
```

### Merge all results

```{r}
## Annotate tables, filter, sort and merge
experiment_1_furin_fes_lfcshrink$gene <- rownames(experiment_1_furin_fes_lfcshrink)
experiment_1_furin_fes_lfcshrink$experiment <- "exp1_sgRNA_06939_FES_FURIN"
experiment_2_znf664_lfcshrink$gene <- rownames(experiment_2_znf664_lfcshrink)
experiment_2_znf664_lfcshrink$experiment <- "exp2_sgRNA_06147_ZNF664"
experiment_3_cnnm2_lfcshrink$gene <- rownames(experiment_3_cnnm2_lfcshrink)
experiment_3_cnnm2_lfcshrink$experiment <- "exp3_sgRNA_25344_CNNM2"
experiment_4_dhx38_lfcshrink$gene <- rownames(experiment_4_dhx38_lfcshrink)
experiment_4_dhx38_lfcshrink$experiment <- "exp4_sgRNA_10966_DHX38"
experiment_5_mat2a_lfcshrink$gene <- rownames(experiment_5_mat2a_lfcshrink)
experiment_5_mat2a_lfcshrink$experiment <- "exp5_sgRNA_02249_MAT2A"

## merge
all_deseq_results <- as.data.frame(rbind(experiment_1_furin_fes_lfcshrink,
                                         experiment_2_znf664_lfcshrink,
                                         experiment_3_cnnm2_lfcshrink,
                                         experiment_4_dhx38_lfcshrink,
                                         experiment_5_mat2a_lfcshrink
                                         ))

## filter
all_deseq_results <- all_deseq_results %>%
  subset(!is.na(padj)) %>%
  arrange(padj)

## significant hits
all_assays_sig <- all_deseq_results %>%
  subset(padj <= 0.05) %>%
  arrange(padj)
```


```{r}
saveRDS(dds,
        file = here("./data/all_validations.deseq2_object.rds"))
```

